---
title: "Using the tidytreatment package"
author: "Joshua J Bon"
date: "`r Sys.Date()`"
bibliography: vignette.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using tidytreatment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette simulates data using the scheme described by @Hill2013 with the additional of 1 categorical variable. It it implemented in the function `simulate_hill_su_data()`:

```{r load-data}

  # before running library(bartMachine), set memory
  options(java.parameters = "-Xmx2000m") # restart R to take effect
  # check memory allocated to Java VM
  options("java.parameters")

  suppressPackageStartupMessages({
    library(bartMachine)
    library(tidytreatment)
    library(dplyr)
    library(tidybayes)
    library(ggplot2)
  })

  sim <- simulate_su_hill_data(n = 200, treatment_linear = F,  omega = 0, add_categorical = T,
                               coef_categorical_treatment = c(0,0,1),  
                               coef_categorical_nontreatment = c(-1,0,-1)
                               )
  
  dat <- sim$data
  dat$c1 <- as.integer(dat$c1)
  # a selection of data
  dat %>% select(y, z, c1, x1:x3) %>% head()

```

## Run the `bartMachine`

Run the model to be used to assess treatment effects. Here we will use `bartMachine`, which is one implementation of Bayesian Additive Regression Trees in `R` [@Kapelner2016]. The package can be found on [CRAN](https://cran.r-project.org/package=bartMachine).
 
```{r run-bart, echo=TRUE, results='hide', cache=TRUE}

  # if you increase the number of cores, the memory needs to be increased,
  # this requires restarting R, setting the 'java.parameters' option then 
  # loading the bartMachine package.
  set_bart_machine_num_cores(2)
  
  # set serialize = T if using the fit over multiple sessions
  # The first bart model will be for the propensity score...
  # i.e. propensity for selection of treatment?
  
  prop_bart <- bartMachine(
    X = select(dat,-y,-z), 
    y = as.factor(select(dat, z)[[1]]), 
    num_burn_in = 2000, 
    num_iterations_after_burn_in = 5000, 
    serialize = T, 
    verbose = F
    ) 

  dat$prop_score <-  prop_bart$p_hat_train
  
  destroy_bart_machine(prop_bart)
  
  prior_incl_prob <- setNames(rep(1, times = ncol(dat) - 1), colnames(dat)[colnames(dat) != "y"])
  
  prior_incl_prob["z"] <- 2
  
  bartM <- bartMachine(
    X = select(dat,-y),
    y = select(dat, y)[[1]],
    num_burn_in = 2000, 
    num_iterations_after_burn_in = 5000,
    serialize = T,
    verbose = F,
    cov_prior_vec = prior_incl_prob
    ) 

  #
```

## Extract the posterior (tidy style)

Methods for extracting the posterior in a tidy format is included in the `tidytreatment`.

```{r tidy-bart, echo=TRUE, cache=TRUE}

  posterior_fitted <- fitted_draws(bartM, value = "fit", include_newdata = F)
  # The newdata argument (omitted) defaults to the data from the model.
  # include_newdata = F, avoids returning the newdata with the fitted values
  # as it is so large. 
  # The `.row` variable makes sure we know which row in the newdata the fitted
  # value came from (if we dont include the data in the result).
  
  posterior_fitted
  
  # Function to tidy predicted draws also...
  posterior_pred <- predicted_draws(bartM, include_newdata = F)
  
  # but this only adds random noise using posterior model variance
  posterior_pred
  
```

## Use some functions from the `tidybayes` package

Since `tidytreatment` follows the `tidybayes` output specifications, functions from `tidybayes` should work.

```{r plot-tidy-bart, echo=TRUE, cache=FALSE}

  treatment_var_and_c1 <- 
    dat %>% 
    select(z,c1) %>%
    mutate(.row = 1:n(), z = as.factor(z))

  posterior_fitted %>%
    left_join(treatment_var_and_c1, by = ".row") %>%
    ggplot() + 
      geom_eye(aes(x = z, y = fit)) + 
      facet_wrap(~c1, labeller = as_labeller( function(x) paste("c1 =",x) ) ) +
      xlab("Treatment (z)") + ylab("Posterior predicted value") +
      theme_bw()

```

## Calculate CATEs

Posterior conditional (average) treatment effects can be calculated using the `posterior_cate_from_sample` function. This function finds the posterior values of
$$
\text{E}(y ~ \vert~ T = 1, X = x_{i}) - \text{E}(y ~ \vert~ T = 0, X = x_{i}) 
$$
for each unit of measurement, $i$, (e.g. subject) in the data sample.

Some histogram summaries are presented below.

```{r cates-hist, echo=TRUE, cache=FALSE}

# sample based (using data from fit) conditional treatment effects, posterior draws
posterior_treat_eff <- 
  treatment_effects(bartM, treatment = "z") # add common support method heer

posterior_treat_eff %>% 
  ggplot() +
  geom_histogram(aes(x = cte), binwidth = 0.1, colour = "white") + 
  theme_bw()

posterior_treat_eff %>% summarise(cte_hat = median(cte)) %>%
  ggplot() +
  geom_histogram(aes(x = cte_hat), binwidth = 0.1, colour = "white") + 
  theme_bw()


```

Plots can be made that stack each subjects posterior CIs of the CATEs. 

```{r cates-stack-plot, echo=TRUE, cache=FALSE}

posterior_treat_eff %>% select(-z) %>% point_interval() %>%
  arrange(cte) %>% mutate(.orow = 1:n()) %>% 
  ggplot() + 
      geom_interval(aes(x = .orow, y= cte), size = 0.5) +
  geom_point(aes(x = .orow, y = cte), shape = "circle open", alpha = 0.1) + 
      ylab("Median posterior CATE for each subject (95% CI)") +
      theme_bw() + coord_flip() + scale_colour_brewer() +
      theme(axis.title.y = element_blank(), 
            axis.text.y = element_blank(), 
            axis.ticks.y = element_blank(),
            legend.position = "none")

csupp1 <- has_common_support(bartM,treatment = "z", method = "chisq", cutoff = 0.05)
csupp1 %>% filter(!common_support)

csupp2 <- has_common_support(bartM,treatment = "z", method = "sd", cutoff = 1)
csupp2 %>% filter(!common_support)

```

We can also plot the CATEs varying over particular covariates. In this example, instead of grouping by subject, we group by the variable of interest, and calculate the posterior summaries over this variable.

```{r cates-line-plot, echo=TRUE, cache=FALSE}

posterior_treat_eff %>%
  left_join(tibble(c1 = dat$c1, .row = 1:length(dat$c1) ), by = ".row") %>%
  group_by(c1, z) %>%
  point_interval(.width = c(0.95)) %>%
  ggplot() + 
      geom_pointinterval(aes(x = c1, y = cte), alpha = 0.2) +
      scale_fill_brewer() +
      facet_wrap(~z) +
      theme_bw()


```

```{r interaction-investigator, echo=T, cache=F}

  treatment_interactions <-
    covariate_with_treatment_importance(bartM, treatment = "z")

  # heterogenity in treatment effect
  treatment_interactions %>% 
    ggplot() + geom_bar(aes(x = variable, y = avg_inclusion), stat = "identity") +
    theme_bw()
  
  
  

```

## References
