---
title: "Benchmark: Vanilla Prediction"
subtitle: "Cross-package comparison of BART-family posterior predictive accuracy, with matched priors"
format:
  html:
    toc: true
    code-fold: true
params:
  fast_mode: true      # tiny hp/n/B for a quick smoke-test render
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(BART)
library(dbarts)
library(stochtree)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/fit-engines-prediction.R")
source("R/metrics.R")
source("R/run-benchmark-prediction.R")

sessioninfo::session_info()
```

## Scope

Engines: `BART::wbart`/`pbart` and `dbarts::bart2` (baseline only - no native
gfr/auto-k/etc. among the settings varied here), `stochtree::bart` (baseline,
`num_gfr = 0`; and `+gfr`, `stochtree`'s own warm-start default). `stan4bart`
is architecturally a mixed-effects+BART model and has no fair "no groups"
mode, so it's excluded here and tested on its own ground in
`benchmark-prediction-rfx.qmd` - `dbarts::bart2` fills that slot here
instead, now that it's a fully supported engine in its own right.
`bartCause` has no vanilla-prediction-only mode (always fits a response *and*
treatment model) and is excluded for the same reason; it's exercised fully in
`benchmark-causal.qmd`. `bartMachine` is excluded throughout per the original
scope decision.

DGP: Friedman's function (Friedman 1991; the field-standard BART benchmark),
continuous and a probit-transformed binary variant - see `R/dgp.R` for the
exact definitions and how "true probability" is defined for the binary case.

## Hyperparameters

```{r hyperparams}
hp <- baseline_hyperparams()
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
str(hp)
```

## Run

```{r run-benchmark}
#| cache: false
n_values <- if (isTRUE(params$fast_mode)) c(30, 60) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps

cache_file <- file.path("cache", sprintf("prediction_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  run_benchmark_prediction(n_values = n_values, B = n_reps, hp = hp, seed = params$seed)
})
```

## Results

### Summary metrics

```{r summary-table}
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(
    across(c(rmse, mae, coverage95, crps, fit_time_sec), mean),
    .groups = "drop"
  ) %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

### Predicted vs. true

One representative replication (largest `n`), posterior mean fitted value
against the known ground truth.

Two separate plots, not one faceted by outcome too: `facet_grid(outcome ~
engine, scales = "free")` gives every individual *cell* its own axis range
(not just each outcome row), which makes engines within the same outcome
visually incomparable - misleading for exactly the side-by-side comparison
this plot exists for.

```{r scatter-plot-continuous}
#| fig-height: 4
res$examples %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Continuous outcome", x = "True value", y = "Posterior mean fitted value") +
  theme_bw()
```

```{r scatter-plot-binary}
#| fig-height: 4
res$examples %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Binary outcome (probability scale)", x = "True probability", y = "Posterior mean fitted probability") +
  theme_bw()
```

### Coverage: nominal vs. empirical

```{r coverage-plot}
#| fig-height: 4
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(coverage95 = mean(coverage95), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = coverage95, colour = factor(n))) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### Cross-engine agreement (baseline rows)

```{r agreement-table}
res$agreement %>%
  group_by(outcome, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), mean_rmse = mean(rmse), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

```{r agreement-heatmap}
#| fig-height: 4
res$agreement %>%
  group_by(outcome, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), .groups = "drop") %>%
  ggplot(aes(x = engine_a, y = engine_b, fill = mean_correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", mean_correlation)), size = 3) +
  facet_wrap(~outcome) +
  scale_fill_distiller(palette = "RdYlBu", direction = 1, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "correlation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Runtime

```{r runtime-plot}
#| fig-height: 4
res$metrics %>%
  group_by(outcome, engine, variant) %>%
  summarise(fit_time_sec = mean(fit_time_sec), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = fit_time_sec, fill = outcome)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Mean fit time (s)") +
  theme_bw()
```

### Flags

Rows whose RMSE is a statistical outlier (|z| > 2) within its own DGP
setting - worth a look, not necessarily a bug (see `flag_outliers()` in
`R/metrics.R` for the important caveat about small-group z-scores).

```{r flags}
flagged <- res$metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(rmse = mean(rmse), .groups = "drop") %>%
  flag_outliers("rmse", c("outcome", "n"))

if (nrow(flagged) == 0) {
  cat("No rows flagged.\n")
} else {
  knitr::kable(flagged, digits = 3)
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
