---
title: "Benchmark: Causal Estimates"
subtitle: "Cross-package comparison of ATE/CATE recovery, with matched priors"
format:
  html:
    toc: true
    code-fold: true
params:
  fast_mode: true
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(BART)
library(stochtree)
library(bartCause)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/metrics.R")
source("R/metrics-causal.R")
source("R/propensity.R")
source("R/fit-engines-causal.R")
source("R/run-benchmark-causal.R")

sessioninfo::session_info()
```

## Scope

This document runs 13 engine/variant/propensity-recipe rows in total (see
`R/run-benchmark-causal.R` for the full list). Three families of causal
engine are compared: `BART` and `stochtree::bart` fit as a two-step
procedure (variable selection, then propensity, then the treatment-effect
model - matching the two-step workflow shown in the package's own BART and
stochtree vignettes), `bartCause::bartc`, which has native causal machinery
including its own internal propensity estimation, and `stochtree::bcf`,
which likewise estimates `tau`/`cate` natively with its own internal
propensity estimation. `stan4bart` is excluded as a standalone causal engine
for the same architectural reason it is excluded from
`benchmark-prediction.qmd`, and `bartMachine` is excluded throughout.

Every "DIY" propensity row - meaning every row where the propensity score is
estimated outside the causal engine itself, rather than natively - is fit
with the same `fit_diy_propensity()` function, shared across all engines
that use it so that the causal engine is the only thing that varies between
rows. Each of these rows is compared under two recipes: `two_stage`, where
variable selection is run first and the propensity model is fit only on the
selected variables (matching the vignettes exactly), and `ps_all`, where the
propensity model instead uses every confounder directly, with no selection
step.

The data-generating process is `simulate_su_hill_data()` (Hill & Su 2013),
used here in a continuous version and in a binary version thresholded at the
median. Each is run under two settings: `response_parallel = TRUE`, which
gives a homogeneous true effect and so is a clean check on ATE recovery, and
`response_parallel = FALSE`, which gives a heterogeneous true effect and so
is where PEHE becomes a meaningful metric. Ground truth for the continuous
outcome comes from `tidytreatment::su_hill_true_effects()`, an exact (not
simulated) reconstruction of the DGP's potential outcomes, verified
bit-for-bit against `simulate_su_hill_data()`'s own `mean_y` in the
package's own test suite. Ground truth for the binary variant, on the
probability scale, comes from this file's own `su_hill_truth_binary()`
(`R/dgp.R`), which is in turn verified against `su_hill_true_effects()` by
direct Monte Carlo.

## Hyperparameters

```{r hyperparams}
hp <- baseline_hyperparams()
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
str(hp)
```

## Run

<details>
<summary>Show run log</summary>

```{r run-benchmark}
n_values <- if (isTRUE(params$fast_mode)) c(40, 60) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps

cache_file <- file.path("cache", sprintf("causal_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  run_benchmark_causal(n_values = n_values, B = n_reps, hp = hp, seed = params$seed)
})
```

</details>

## Results

### Summary metrics

```{r summary-table}
res$metrics %>%
  group_by(outcome, response_parallel, engine, variant, propensity_recipe, n) %>%
  summarise(
    across(c(ate_bias, ate_covered, att_bias, att_covered, pehe, fit_time_sec), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(outcome, response_parallel, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

### CATE-hat vs. true CATE

These plots show one representative replication, the one with the largest
`n`, and only the heterogeneous-truth DGP setting (`response_parallel =
FALSE`) - the homogeneous setting (`TRUE`) has a constant true effect, so
there is no relationship to show. As with the equivalent plot in
`benchmark-prediction.qmd`, the continuous and binary outcomes get two
separate plots rather than one faceted by outcome, so that axes stay shared
within an outcome instead of floating freely per panel, keeping the engines
visually comparable.

```{r cate-scatter-continuous}
#| fig-height: 7
res$examples %>%
  filter(outcome == "continuous") %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = true_ite, y = cte_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~row_label, ncol = 4) +
  labs(title = "Continuous outcome", x = "True CATE", y = "Posterior mean CATE-hat") +
  theme_bw()
```

```{r cate-scatter-binary}
#| fig-height: 7
res$examples %>%
  filter(outcome == "binary") %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = true_ite, y = cte_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~row_label, ncol = 4) +
  labs(title = "Binary outcome (probability scale)", x = "True CATE (probability scale)", y = "Posterior mean CATE-hat") +
  theme_bw()
```

### ATE / ATT coverage: nominal vs. empirical

```{r coverage-plot}
#| fig-height: 6
res$metrics %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, response_parallel, row_label, n) %>%
  summarise(across(c(ate_covered, att_covered), mean), .groups = "drop") %>%
  pivot_longer(c(ate_covered, att_covered), names_to = "estimand", values_to = "coverage") %>%
  ggplot(aes(x = row_label, y = coverage, colour = factor(n), shape = estimand)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_grid(response_parallel ~ outcome, labeller = label_both) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### PEHE (heterogeneous-truth setting only)

```{r pehe-plot}
#| fig-height: 5
res$metrics %>%
  filter(!is.na(pehe)) %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, row_label, n) %>%
  summarise(pehe = mean(pehe), .groups = "drop") %>%
  ggplot(aes(x = row_label, y = pehe, fill = factor(n))) +
  geom_col(position = "dodge") +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "PEHE", fill = "n") +
  theme_bw()
```

### Cross-engine CATE agreement (baseline rows, by propensity recipe)

```{r agreement-table}
res$agreement %>%
  group_by(outcome, response_parallel, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), mean_rmse = mean(rmse), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

```{r agreement-heatmap}
#| fig-height: 10
#| fig-width: 9
res$agreement %>%
  group_by(outcome, response_parallel, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), .groups = "drop") %>%
  ggplot(aes(x = engine_a, y = engine_b, fill = mean_correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", mean_correlation)), size = 2.5) +
  facet_grid(response_parallel ~ outcome, labeller = label_both) +
  scale_fill_distiller(palette = "RdYlBu", direction = 1, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "correlation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Runtime

```{r runtime-plot}
#| fig-height: 6
res$metrics %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, row_label) %>%
  summarise(fit_time_sec = mean(fit_time_sec), .groups = "drop") %>%
  ggplot(aes(x = row_label, y = fit_time_sec, fill = outcome)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Mean fit time (s)") +
  theme_bw()
```

### Common support: tidytreatment vs. bartCause's native `commonSup.rule`

This table reports the agreement rate between bartCause's own native
SD-rule common-support flag and tidytreatment's independently-coded
`has_common_support()`, both applied to the *same* fitted response-stage
model (`bartc_fit$fit.rsp`). See `bartc_common_support_agreement()` in
`R/metrics-causal.R` for how the comparison is made.

```{r common-support-table}
res$common_support %>%
  group_by(outcome, response_parallel, variant, propensity_recipe) %>%
  summarise(mean_agreement_rate = mean(agreement_rate), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

### Flags

This section lists rows whose PEHE (in the heterogeneous-truth setting) or
ATE bias (in the homogeneous setting) is a statistical outlier (|z| > 2)
within its own DGP setting. A flagged row is worth a closer look, but is not
necessarily a bug - see `flag_outliers()` in `R/metrics.R` for the caveat
about small-group z-scores.

```{r flags}
flagged_pehe <- res$metrics %>%
  filter(!is.na(pehe)) %>%
  group_by(outcome, n, engine, variant, propensity_recipe) %>%
  summarise(pehe = mean(pehe), .groups = "drop") %>%
  flag_outliers("pehe", c("outcome", "n"))

flagged_ate <- res$metrics %>%
  group_by(outcome, n, engine, variant, propensity_recipe) %>%
  summarise(abs_ate_bias = mean(abs(ate_bias)), .groups = "drop") %>%
  flag_outliers("abs_ate_bias", c("outcome", "n"))

if (nrow(flagged_pehe) == 0) {
  cat("No PEHE outliers flagged.\n")
} else {
  knitr::kable(flagged_pehe, digits = 3, caption = "PEHE outliers")
}

if (nrow(flagged_ate) == 0) {
  cat("No |ATE bias| outliers flagged.\n")
} else {
  knitr::kable(flagged_ate, digits = 3, caption = "|ATE bias| outliers")
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
