---
title: "Benchmark: Prediction With Random Effects"
subtitle: "stan4bart vs. stochtree::bart(rfx_group_ids), on ground both were built for"
format:
  html:
    toc: true
    code-fold: true
params:
  fast_mode: true
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(stan4bart)
library(stochtree)
library(dbarts)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/metrics.R")
source("R/fit-engines-rfx.R")
source("R/metrics-rfx.R")
source("R/run-benchmark-rfx.R")

sessioninfo::session_info()
```

## Scope

`benchmark-prediction.qmd` excludes `stan4bart` entirely: it's architecturally
a mixed-effects+BART model (its formula parser requires at least one
`(1|group)` term), so it has no fair "no groups" mode to compare against a
plain BART fit. This document tests it on the ground it was actually built
for instead, against `stochtree::bart()` with `rfx_group_ids_train`
(intercept-only random effects - the direct analogue of `(1|group)`).
`BART`/`bartCause`/`stochtree::bcf` have no random-effects prediction mode
and are excluded. `bartMachine` is excluded throughout.

4 rows: `stan4bart` baseline/`+auto_k`, `stochtree::bart` baseline/`+gfr`.

DGP: `simulate_friedman_rfx()`/`simulate_friedman_rfx_binary()` (`R/dgp.R`) -
Friedman's function plus an explicit random intercept per group, with exact
ground truth for both the fixed-effect surface and the true per-group
effects (unlike `simulate_su_hill_data()`'s own `n_subjects`/`sd_subjects`
mechanism, which adds the random effect to noisy `y` directly and doesn't
return the true effects at all).

Beyond the usual prediction-accuracy metrics, this document's sharpest
diagnostic is **group-effect recovery**: does each engine's own posterior
mean per-group random intercept actually track the true simulated group
effects? Note BART's trees can absorb some of the overall baseline level
that would otherwise land in the random intercepts (the two aren't uniquely
identified from data alone), so recovery is reported as correlation plus
*centred* RMSE (shared shift removed) rather than raw bias.

## Hyperparameters

```{r hyperparams}
hp <- baseline_hyperparams()
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
str(hp)
```

## Run

```{r run-benchmark}
n_values <- if (isTRUE(params$fast_mode)) c(60, 100) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps

cache_file <- file.path("cache", sprintf("rfx_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  run_benchmark_rfx(n_values = n_values, B = n_reps, hp = hp, seed = params$seed)
})
```

## Results

### Prediction accuracy

```{r summary-table}
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(across(c(rmse, mae, coverage95, crps, fit_time_sec), mean), .groups = "drop") %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

### Predicted vs. true (fixed-effect surface + group effect combined)

One representative replication (largest `n`).

```{r pred-scatter-continuous}
#| fig-height: 4
res$examples %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Continuous outcome", x = "True value (f(x) + group effect)", y = "Posterior mean fitted value") +
  theme_bw()
```

```{r pred-scatter-binary}
#| fig-height: 4
res$examples %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Binary outcome (probability scale)", x = "True probability", y = "Posterior mean fitted probability") +
  theme_bw()
```

### Group-effect recovery

The headline diagnostic for this document: does each engine's own posterior
mean per-group random intercept track the true simulated group effects?
Centred (see caption) values are comparable in scale across engines within
the same outcome, so each outcome gets one plot with shared (not free) axes
across its engine panels - `facet_grid(..., scales = "free")` across the
engine dimension would give each panel its own axis range independently,
which is misleading for exactly this side-by-side comparison (see the same
issue, diagnosed in detail, for `benchmark-prediction.qmd`'s predicted-vs-true
plot).

```{r group-effect-table}
res$group_metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(across(c(group_effect_cor, group_effect_centered_rmse), mean), .groups = "drop") %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

```{r group-effect-scatter-prep}
# Centred, matching group_effect_recovery()'s own metric definition: the
# shared level between BART's trees and the random intercept isn't
# identified, so only the *relative* effects are comparable.
group_examples_centred <- res$group_examples %>%
  group_by(outcome, engine, variant) %>%
  mutate(
    true_centred = true_group_effect - mean(true_group_effect),
    hat_centred = group_effect_hat - mean(group_effect_hat)
  ) %>%
  ungroup()
```

```{r group-effect-scatter-continuous}
#| fig-height: 4
group_examples_centred %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = true_centred, y = hat_centred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 1.5) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Continuous outcome", x = "True group effect (centred)", y = "Posterior mean group effect estimate (centred)") +
  theme_bw()
```

```{r group-effect-scatter-binary}
#| fig-height: 4
group_examples_centred %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = true_centred, y = hat_centred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 1.5) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Binary outcome (probit latent scale)", x = "True group effect (centred)", y = "Posterior mean group effect estimate (centred)") +
  theme_bw()
```

### Coverage: nominal vs. empirical

```{r coverage-plot}
#| fig-height: 4
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(coverage95 = mean(coverage95), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = coverage95, colour = factor(n))) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### Runtime

```{r runtime-plot}
#| fig-height: 3
res$metrics %>%
  group_by(outcome, engine, variant) %>%
  summarise(fit_time_sec = mean(fit_time_sec), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = fit_time_sec, fill = outcome)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Mean fit time (s)") +
  theme_bw()
```

### Flags

Rows whose RMSE (prediction) or centred group-effect RMSE (recovery) is a
statistical outlier (|z| > 2) within its own outcome/n setting (see
`flag_outliers()` in `R/metrics.R` for the small-group z-score caveat - with
only 4 rows per group here, this is a coarse screen).

```{r flags}
flagged_rmse <- res$metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(rmse = mean(rmse), .groups = "drop") %>%
  flag_outliers("rmse", c("outcome", "n"))

flagged_group <- res$group_metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(group_effect_centered_rmse = mean(group_effect_centered_rmse), .groups = "drop") %>%
  flag_outliers("group_effect_centered_rmse", c("outcome", "n"))

if (nrow(flagged_rmse) == 0) {
  cat("No prediction RMSE outliers flagged.\n")
} else {
  knitr::kable(flagged_rmse, digits = 3, caption = "Prediction RMSE outliers")
}

if (nrow(flagged_group) == 0) {
  cat("No group-effect-recovery outliers flagged.\n")
} else {
  knitr::kable(flagged_group, digits = 3, caption = "Group-effect recovery outliers")
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
