---
title: "Benchmark: Causal Estimates"
subtitle: "Cross-package comparison of ATE/CATE recovery, with matched priors"
format:
  html:
    toc: true
    code-fold: true
params:
  fast_mode: true
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(BART)
library(stochtree)
library(bartCause)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/metrics.R")
source("R/metrics-causal.R")
source("R/propensity.R")
source("R/fit-engines-causal.R")
source("R/run-benchmark-causal.R")

sessioninfo::session_info()
```

## Scope

13 engine/variant/propensity-recipe rows (see `R/run-benchmark-causal.R`):
`BART`/`stochtree::bart` two-step (VS -> PS -> TE, matching the package's own
BART/stochtree vignettes), `bartCause::bartc` (native causal machinery,
including its own internal propensity estimation), and `stochtree::bcf`
(native `tau`/`cate`, including its own internal propensity estimation).
`stan4bart` is excluded as a standalone causal engine for the same
architectural reason it's excluded from `benchmark-prediction.qmd`.
`bartMachine` is excluded throughout.

Every "DIY" propensity row is fit with `fit_diy_propensity()` (shared across
all engines that use it, so the causal engine is the only thing varying) and
compared under two recipes: `two_stage` (variable-selection then propensity,
matching the vignettes exactly) and `ps_all` (propensity on every confounder
directly, no selection).

DGP: `simulate_su_hill_data()` (Hill & Su 2013), continuous and a
median-thresholded binary variant, each run under `response_parallel = TRUE`
(homogeneous truth - clean ATE recovery check) and `FALSE` (heterogeneous
truth - PEHE becomes meaningful). Ground truth via
`tidytreatment::su_hill_true_effects()` (an exact, not simulated,
reconstruction of the DGP's potential outcomes - verified bit-for-bit against
`simulate_su_hill_data()`'s own `mean_y` in the package's own test suite) and
this file's own `su_hill_truth_binary()` (`R/dgp.R`) for the binary variant's
probability-scale truth, verified against it by direct Monte Carlo.

## Hyperparameters

```{r hyperparams}
hp <- baseline_hyperparams()
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
str(hp)
```

## Run

```{r run-benchmark}
n_values <- if (isTRUE(params$fast_mode)) c(40, 60) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps

cache_file <- file.path("cache", sprintf("causal_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  run_benchmark_causal(n_values = n_values, B = n_reps, hp = hp, seed = params$seed)
})
```

## Results

### Summary metrics

```{r summary-table}
res$metrics %>%
  group_by(outcome, response_parallel, engine, variant, propensity_recipe, n) %>%
  summarise(
    across(c(ate_bias, ate_covered, att_bias, att_covered, pehe, fit_time_sec), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(outcome, response_parallel, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

### CATE-hat vs. true CATE

One representative replication (largest `n`, heterogeneous-truth DGP setting
only - `response_parallel = FALSE`, since `TRUE` has a constant true effect
with nothing to show a relationship against). Two separate plots (not one
faceted by outcome) for the same reason as the equivalent plot in
`benchmark-prediction.qmd`: shared axes within an outcome, not per-panel free
scales, so engines are visually comparable.

```{r cate-scatter-continuous}
#| fig-height: 7
res$examples %>%
  filter(outcome == "continuous") %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = true_ite, y = cte_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~row_label, ncol = 4) +
  labs(title = "Continuous outcome", x = "True CATE", y = "Posterior mean CATE-hat") +
  theme_bw()
```

```{r cate-scatter-binary}
#| fig-height: 7
res$examples %>%
  filter(outcome == "binary") %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = true_ite, y = cte_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~row_label, ncol = 4) +
  labs(title = "Binary outcome (probability scale)", x = "True CATE (probability scale)", y = "Posterior mean CATE-hat") +
  theme_bw()
```

### ATE / ATT coverage: nominal vs. empirical

```{r coverage-plot}
#| fig-height: 6
res$metrics %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, response_parallel, row_label, n) %>%
  summarise(across(c(ate_covered, att_covered), mean), .groups = "drop") %>%
  pivot_longer(c(ate_covered, att_covered), names_to = "estimand", values_to = "coverage") %>%
  ggplot(aes(x = row_label, y = coverage, colour = factor(n), shape = estimand)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_grid(response_parallel ~ outcome, labeller = label_both) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### PEHE (heterogeneous-truth setting only)

```{r pehe-plot}
#| fig-height: 5
res$metrics %>%
  filter(!is.na(pehe)) %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, row_label, n) %>%
  summarise(pehe = mean(pehe), .groups = "drop") %>%
  ggplot(aes(x = row_label, y = pehe, fill = factor(n))) +
  geom_col(position = "dodge") +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "PEHE", fill = "n") +
  theme_bw()
```

### Cross-engine CATE agreement (baseline rows, by propensity recipe)

```{r agreement-table}
res$agreement %>%
  group_by(outcome, response_parallel, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), mean_rmse = mean(rmse), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

```{r agreement-heatmap}
#| fig-height: 10
#| fig-width: 9
res$agreement %>%
  group_by(outcome, response_parallel, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), .groups = "drop") %>%
  ggplot(aes(x = engine_a, y = engine_b, fill = mean_correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", mean_correlation)), size = 2.5) +
  facet_grid(response_parallel ~ outcome, labeller = label_both) +
  scale_fill_distiller(palette = "RdYlBu", direction = 1, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "correlation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Runtime

```{r runtime-plot}
#| fig-height: 6
res$metrics %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, row_label) %>%
  summarise(fit_time_sec = mean(fit_time_sec), .groups = "drop") %>%
  ggplot(aes(x = row_label, y = fit_time_sec, fill = outcome)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Mean fit time (s)") +
  theme_bw()
```

### Common support: tidytreatment vs. bartCause's native `commonSup.rule`

Agreement rate between bartCause's own native SD-rule common-support flag
and tidytreatment's independently-coded `has_common_support()`, applied to
the *same* fitted response-stage model (`bartc_fit$fit.rsp`) - see
`bartc_common_support_agreement()` in `R/metrics-causal.R`.

```{r common-support-table}
res$common_support %>%
  group_by(outcome, response_parallel, variant, propensity_recipe) %>%
  summarise(mean_agreement_rate = mean(agreement_rate), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

### Flags

Rows whose PEHE (heterogeneous-truth setting) or ATE bias (homogeneous
setting) is a statistical outlier (|z| > 2) within its own DGP setting -
worth a look, not necessarily a bug (see `flag_outliers()` in `R/metrics.R`
for the small-group z-score caveat).

```{r flags}
flagged_pehe <- res$metrics %>%
  filter(!is.na(pehe)) %>%
  group_by(outcome, n, engine, variant, propensity_recipe) %>%
  summarise(pehe = mean(pehe), .groups = "drop") %>%
  flag_outliers("pehe", c("outcome", "n"))

flagged_ate <- res$metrics %>%
  group_by(outcome, n, engine, variant, propensity_recipe) %>%
  summarise(abs_ate_bias = mean(abs(ate_bias)), .groups = "drop") %>%
  flag_outliers("abs_ate_bias", c("outcome", "n"))

if (nrow(flagged_pehe) == 0) {
  cat("No PEHE outliers flagged.\n")
} else {
  knitr::kable(flagged_pehe, digits = 3, caption = "PEHE outliers")
}

if (nrow(flagged_ate) == 0) {
  cat("No |ATE bias| outliers flagged.\n")
} else {
  knitr::kable(flagged_ate, digits = 3, caption = "|ATE bias| outliers")
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
