---
title: "Benchmark: Causal Estimates"
subtitle: "Cross-package comparison of ATE/CATE recovery, with matched priors"
format:
  html:
    toc: true
    code-fold: true
    html-math-method: katex
bibliography: refs.bib
params:
  fast_mode: true
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
  y_sd: 1               # SD of the noise added to the continuous outcome
  parallel_workers: null # number of parallel workers for replications; null -> availableCores() - 1, 1 disables parallelism
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(BART)
library(stochtree)
library(bartCause)
library(future)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/metrics.R")
source("R/metrics-causal.R")
source("R/propensity.R")
source("R/fit-engines-causal.R")
source("R/parallel-driver.R")
source("R/run-benchmark-causal.R")

benchmark_dir <- normalizePath(".")

```

## Scope

This document compares causal-effect estimation across three families of
engine:

- `BART`/`stochtree::bart` fit as a two-step procedure - variable selection,
  then propensity, then the treatment-effect model - matching each
  package's own two-step vignette.
- `bartCause::bartc`, with native causal machinery including its own
  propensity estimation.
- `stochtree::bcf`, which estimates `tau`/`cate` natively with its own
  propensity estimation.

`stan4bart` is excluded here for the same architectural reason as in
`benchmark-prediction.qmd`, and `bartMachine` is excluded throughout. Each
engine also gets a `default` row, using its own out-of-the-box hyperprior
settings instead of the matched `baseline` - see Package-default hyperprior
variants below (see `R/run-benchmark-causal.R` for the full row list, which
now differs slightly by outcome).

Every "DIY" propensity row - where the propensity score is estimated outside
the causal engine itself - is fit with the same `fit_diy_propensity()`
function, so the causal engine is the only thing that varies between rows.
Each such row is compared under two recipes:

- `two_stage`: variable selection first, propensity fit only on the
  selected variables (matching the vignettes).
- `ps_all`: propensity fit on every confounder directly, with no selection
  step.

## Data generating process

The data-generating process is `simulate_su_hill_data()` [@hillsu2013],
used in a continuous version and a binary version thresholded at the
median. Each runs under two settings: `response_parallel = TRUE`
(homogeneous true effect, a clean ATE check) and `response_parallel =
FALSE` (heterogeneous true effect, where PEHE becomes meaningful). Ground
truth for the continuous outcome comes from
`tidytreatment::su_hill_true_effects()`, verified bit-for-bit against the
package's own test suite. Ground truth for the binary variant comes from
this file's own `su_hill_truth_binary()` (`R/dgp.R`), verified against
`su_hill_true_effects()` by direct Monte Carlo.

For the continuous outcome, independent $N(0, \sigma^2)$ noise is added on top of `simulate_su_hill_data()`'s own mean response^[By default $\sigma = 1$ but this quarto file contains the `y_sd` document parameter to vary this.].

### Estimated signal-to-noise ratio (continuous case)

```{r snr-demo}
set.seed(1)
n_snr <- 5000
sim_snr_par <- simulate_su_hill_data(n = n_snr, tau = 4, response_parallel = TRUE, y_sd = params$y_sd)
sim_snr_het <- simulate_su_hill_data(n = n_snr, tau = 4, response_parallel = FALSE, y_sd = params$y_sd)
snr_par <- var(sim_snr_par$mean_y) / params$y_sd^2
snr_het <- var(sim_snr_het$mean_y) / params$y_sd^2
```

The global variance is $\sigma^2 = `r params$y_sd^2`$, and the variance ratio^[The variance of the mean response, relative to $\sigma^2$.] is estimated to be `r round(snr_par, 2)` for the homogeneous-effect setting and `r round(snr_het, 2)` for the heterogeneous-effect setting, or **`r round(sqrt(snr_par), 2)`**/**`r round(sqrt(snr_het), 2)`** on the standardised scale (square root).

## Hyperparameters

The main hyperparameters are specified below.

```{r hyperparams}
hp <- baseline_hyperparams()
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
hp_vals <- unlist(hp)
hp_display <- ifelse(hp_vals == round(hp_vals), sprintf("%.0f", hp_vals), as.character(hp_vals))
names(hp_display) <- sprintf("`%s`", names(hp))
knitr::kable(t(hp_display), align = "c", row.names = FALSE)
```

For the `baseline` variant, most of the hyperprior parameter specification
is trivial to match across packages. However, some hyperpriors are chosen
based on the data itself (more or less an empirical Bayes decision) - see
`R/hyperparams.R` for the exact translation used by each engine, including
`stochtree::bcf`'s separate prognostic/treatment-effect forests.

### Tree-structure prior

Naming conventions for the prior on trees differ across packages: `alpha` =
`base` and `beta`=`power`.

These control the prior probability that a node at depth $d$ splits
further: $$P(\text{split at depth } d) = \alpha (1+d)^{-\beta}.$$
These are constants, passed unchanged to whichever argument name each
package uses.

### Leaf-shrinkage prior

Each leaf's contribution gets a $N(0, \tau^2)$ prior, with $\tau$ set from
`k` (the number of prior standard deviations a leaf is expected to sit from
the boundary of $y$'s range) and the number of trees:

$$
\tau = \begin{cases}
\dfrac{\max(y) - \min(y)}{2k\sqrt{\texttt{n\_trees}}} & \text{continuous outcome - uses the data's own range} \\[8pt]
\dfrac{3}{k\sqrt{\texttt{n\_trees}}} & \text{binary outcome - calibrated to the probit latent's } \pm 3 \text{ range instead}
\end{cases}
$$

Some differences between packages:

- `BART`/`dbarts`-style engines (including `bartCause`, which shares
  `dbarts::bart2`'s argument set) take $\tau$ via `k` directly.
- `stochtree` standardizes a continuous outcome internally, so it needs the
  *variance* $\tau^2$ on that standardized scale instead. This is $\tau^2 /
  \text{var}(y)$ for a continuous outcome, or plain $\tau^2$ for binary (the
  probit latent isn't rescaled). `stochtree::bcf` applies this to both its
  prognostic and treatment-effect forests.

More generally, `BART::wbart` and `stochtree` (both `bart` and `bcf`) centre the response before fitting for a continuous outcome (`stochtree` also rescales it, per the standardisation above), while `dbarts::bart2` - and so `bartCause`, which shares it - works on the raw scale throughout. Neither centering concept applies to the binary outcome, fit directly on the 0/1 response under a probit link in every package here.

### Global error-variance prior

:::{.callout-note}
The following applies to the real-valued response only.
:::

The residual variance $\sigma^2$ has an inverse-gamma prior,
$$\sigma^2 \sim \text{Inv-Gamma}(\alpha, \beta),$$
with shape $\alpha = \nu / 2$ set from $\nu =$ `sigdf` degrees of freedom.

An intermediate quantity $\lambda$ is calibrated from the data so that the
prior places `sigquant` probability below a rough, data-driven estimate of
$\sigma$ itself, giving the scale $\beta$:

$$
\beta = \frac{\nu \lambda}{2}, \qquad
\lambda = \hat\sigma^2 \cdot \frac{\chi^2_{\nu,\,1 - \texttt{sigquant}}}{\nu}, \qquad
\hat\sigma =  \begin{cases}
  \hat\sigma_\text{LM-MLE} & \text{if}~n > p,\\
  \hat\sigma_y & \text{otherwise},
\end{cases}
$$
where $\chi^2_{\nu,\,p}$ is the $p$-quantile of a $\chi^2_\nu$ distribution, where $\hat\sigma_\text{LM-MLE}$ is the residual SD of a linear fit regressing the response against the covariates, and where $\hat\sigma_y$ is the standard deviation of the data $y$.

`BART`/`dbarts`-style engines compute $\hat\sigma$, $\lambda$, and this
inverse-gamma prior internally when `sigest` is left at its default.
`stochtree` has no equivalent auto-calibration of its own - matching it
means passing `stochtree`'s own `shape`/`scale` arguments directly, on its
internally-standardized scale:

$$
\text{shape} = \alpha = \frac{\nu}{2}, \qquad \text{scale} = \beta_{\text{std}} = \frac{\nu \lambda}{2 \text{var}(y)}.
$$

## Package-default hyperprior variants

Every engine also gets a `default` row, using its own out-of-the-box
hyperprior settings instead, while burn-in, posterior draws, number of
chains, and `stochtree`'s warm-start/adaptive-coding settings stay fixed at
`baseline`'s values - those are separate axes already varied by other means
(`+auto_k`, `+gfr_adaptive_coding`), not hyperprior settings.

- **`BART` two-step**: no `default` row for the continuous outcome -
  `BART::wbart`'s own defaults already equal `baseline` (see
  `benchmark-prediction.qmd`). The binary outcome uses `BART::pbart`, whose
  `ntree` default (50 vs 200) does differ.
- **`stochtree::bart` two-step**: same as `stochtree::bart`'s `default` row
  in `benchmark-prediction.qmd` - leaf variance reverts to a sampled,
  package-default value, and (continuous outcome only) the global
  error-variance prior reverts to the improper `shape = scale = 0`.
- **`bartc`**: its response/treatment models are genuine `dbarts::bart2()`
  calls (see Global error-variance prior above), so `default` here is the
  same `dbarts::bart2()` package defaults as `benchmark-prediction.qmd`'s
  `dbarts::bart2` `default` row - `n.trees = 75` (vs 200) and `k = NULL`
  (auto-tuned, vs fixed at 2) for both the response and treatment model.
- **`stochtree::bcf`**: a bigger deviation than `stochtree::bart` - `bcf()`'s
  own `num_trees`/`alpha`/`beta` defaults are asymmetric between its two
  forests (250/0.95/2 prognostic, 100/0.25/3 treatment-effect), where
  `baseline` explicitly overrides both to the same shared values (see
  `R/hyperparams.R`). `default` reverts both forests to their own separate
  defaults, on top of the same leaf-variance and global-error-variance
  reversions as `stochtree::bart`.

## Run

<details>
<summary>Show run log</summary>

```{r run-benchmark}
n_values <- if (isTRUE(params$fast_mode)) c(40, 60) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps
n_for_examples <- max(n_values)

cache_file <- file.path("cache", sprintf("causal_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  old_plan <- future::plan(future::multisession, workers = benchmark_worker_count(params$parallel_workers))
  on.exit(future::plan(old_plan), add = TRUE)
  run_benchmark_causal(n_values = n_values, B = n_reps, hp = hp, seed = params$seed, y_sd = params$y_sd,
                        benchmark_dir = benchmark_dir)
})
```

</details>

## Results

### Summary metrics

`fit_time_sec` is reported as a median, not a mean - a handful of runs (see
@sec-runtime) are extreme outliers.

```{r summary-table}
res$metrics %>%
  group_by(outcome, response_parallel, engine, variant, propensity_recipe, n) %>%
  summarise(
    across(c(ate_bias, ate_covered, att_bias, att_covered, pehe), ~ mean(.x, na.rm = TRUE)),
    fit_time_sec = median(fit_time_sec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(outcome, response_parallel, n, engine, variant, propensity_recipe,
         ate_bias, ate_covered, att_bias, att_covered, pehe, fit_time_sec) %>%
  arrange(outcome, response_parallel, n, engine, variant) %>%
  knitr::kable(digits = 3, format = "html") %>%
  kableExtra::collapse_rows(columns = 1:3, valign = "top")
```

### CATE-hat vs. true CATE

These plots show one representative replication for `` n = `r n_for_examples` ``
(the largest value tested), heterogeneous-truth setting only
(`response_parallel = FALSE`) - the homogeneous setting has a constant true
effect, so there's nothing to plot. Continuous and binary outcomes get
separate plots so axes stay shared within each outcome.

```{r cate-scatter-continuous}
#| fig-height: 7
res$examples %>%
  filter(outcome == "continuous") %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = true_ite, y = cte_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~row_label, ncol = 4) +
  labs(title = "Continuous outcome", x = "True CATE", y = "Posterior mean CATE-hat") +
  theme_bw()
```

```{r cate-scatter-binary}
#| fig-height: 7
res$examples %>%
  filter(outcome == "binary") %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = true_ite, y = cte_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~row_label, ncol = 4) +
  labs(title = "Binary outcome (probability scale)", x = "True CATE (probability scale)", y = "Posterior mean CATE-hat") +
  theme_bw()
```

### ATE / ATT coverage: nominal vs. empirical

```{r coverage-plot}
#| fig-height: 6
res$metrics %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, response_parallel, row_label, n) %>%
  summarise(across(c(ate_covered, att_covered), mean), .groups = "drop") %>%
  pivot_longer(c(ate_covered, att_covered), names_to = "estimand", values_to = "coverage") %>%
  ggplot(aes(x = row_label, y = coverage, colour = factor(n), shape = estimand)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_grid(response_parallel ~ outcome, labeller = label_both) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### PEHE (heterogeneous-truth setting only)

```{r pehe-plot}
#| fig-height: 5
res$metrics %>%
  filter(!is.na(pehe)) %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  group_by(outcome, row_label, n) %>%
  summarise(pehe = mean(pehe), .groups = "drop") %>%
  ggplot(aes(x = row_label, y = pehe, fill = factor(n))) +
  geom_col(position = "dodge") +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "PEHE", fill = "n") +
  theme_bw()
```

### Cross-engine CATE agreement (baseline rows, by propensity recipe)

```{r agreement-table}
res$agreement %>%
  group_by(outcome, response_parallel, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), mean_rmse = mean(rmse), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

```{r agreement-heatmap}
#| fig-height: 10
#| fig-width: 9
res$agreement %>%
  group_by(outcome, response_parallel, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), .groups = "drop") %>%
  ggplot(aes(x = engine_a, y = engine_b, fill = mean_correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", mean_correlation)), size = 2.5) +
  facet_grid(response_parallel ~ outcome, labeller = label_both) +
  scale_fill_distiller(palette = "RdYlBu", direction = 1, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "correlation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Runtime {#sec-runtime}

The box plot below shows the full per-replication fit-time distribution on a
log scale, which is where the outliers actually show up - as individual
points well above the whiskers, rather than as a shift in a single summary
number.

```{r runtime-boxplot}
#| fig-height: 8
res$metrics %>%
  mutate(row_label = paste(engine, variant, propensity_recipe)) %>%
  ggplot(aes(x = row_label, y = fit_time_sec, colour = factor(n))) +
  geom_boxplot(outlier.alpha = 0.5) +
  facet_wrap(~outcome) +
  coord_flip() +
  scale_y_log10() +
  labs(x = NULL, y = "Fit time (s, log scale)", colour = "n") +
  theme_bw()
```

### Common support: tidytreatment vs. bartCause's native `commonSup.rule`

This table reports the agreement rate between bartCause's own native
SD-rule common-support flag and tidytreatment's independently-coded
`has_common_support()`, both applied to the *same* fitted response-stage
model (`bartc_fit$fit.rsp`). See `bartc_common_support_agreement()` in
`R/metrics-causal.R` for how the comparison is made.

```{r common-support-table}
res$common_support %>%
  group_by(outcome, response_parallel, variant, propensity_recipe) %>%
  summarise(mean_agreement_rate = mean(agreement_rate), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

### Flags

This section lists rows whose PEHE (in the heterogeneous-truth setting) or
ATE bias (in the homogeneous setting) is a statistical outlier (|z| > 2)
within its own DGP setting. A flagged row is worth a closer look, but is not
necessarily a bug - see `flag_outliers()` in `R/metrics.R` for the caveat
about small-group z-scores.

```{r flags}
flagged_pehe <- res$metrics %>%
  filter(!is.na(pehe)) %>%
  group_by(outcome, n, engine, variant, propensity_recipe) %>%
  summarise(pehe = mean(pehe), .groups = "drop") %>%
  flag_outliers("pehe", c("outcome", "n"))

flagged_ate <- res$metrics %>%
  group_by(outcome, n, engine, variant, propensity_recipe) %>%
  summarise(abs_ate_bias = mean(abs(ate_bias)), .groups = "drop") %>%
  flag_outliers("abs_ate_bias", c("outcome", "n"))

if (nrow(flagged_pehe) == 0) {
  cat("No PEHE outliers flagged.\n")
} else {
  knitr::kable(flagged_pehe, digits = 3, caption = "PEHE outliers")
}

if (nrow(flagged_ate) == 0) {
  cat("No |ATE bias| outliers flagged.\n")
} else {
  knitr::kable(flagged_ate, digits = 3, caption = "|ATE bias| outliers")
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
