---
title: "Benchmark: Prediction With Random Effects"
subtitle: "Cross-package comparison of BART-family posterior predictive accuracy"
format:
  html:
    toc: true
    code-fold: true
    html-math-method: katex
bibliography: refs.bib
params:
  fast_mode: false
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
  y_sd: 1               # SD of the noise added to the continuous outcome
  sd_group_continuous: 3 # SD of the random group intercepts, continuous outcome
  sd_group_binary: 0.5   # SD of the random group intercepts, binary outcome (probit latent scale)
  n_groups: 25            # number of random-effect groups
  n_mcmc_draws: 2000                # number of MCMC draws retained per chain, post burn-in/pre-thinning
  mcmc_seed: 17                      # random seed for the samplers (stan4bart/stochtree), distinct from `seed` (the DGP's own)
  parallel_workers: null           # number of parallel workers for replications; null -> availableCores() - 1, 1 disables parallelism
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(stan4bart)
library(stochtree)
library(dbarts)
library(coda)
library(GGally)
library(future)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/metrics.R")
source("R/fit-engines-rfx.R")
source("R/metrics-rfx.R")
source("R/parallel-driver.R")
source("R/run-benchmark-rfx.R")

benchmark_dir <- normalizePath(".")

```

## Scope

This document compares predictions from two Bayesian Additive Regression Tree (BART) packages in R that can handle the addition of group-specific random effects. The packages tested are `stan4bart` and `stochtree`.

A `baseline` variant is implemented for engines `stan4bart::stan4bart` and `stochtree::bart`. These attempt to match hyperprior specification across the methods, for a fair comparison. Both engines also get a `default` variant, using their own out-of-the-box priors instead. Note that the global variance term for `stan4bart` does not have an inverse gamma prior available, unlike most other BART packages, but does have several other choices (see @sec-prior-global-var for more details).

The method `stochtree::bart` gets one additional variant:

- `baseline+gfr`: using its own warm-start default and otherwise identical to `baseline`.

`stan4bart::stan4bart` gets one additional variant:

- `default+sigest`: (for continuous only) keeps `default`'s prior *family* but calibrates its rate from a potentially more indicative variance estimate (see @sec-prior-global-var for details).

The final number of retained draws is `T=`r params$n_mcmc_draws`` in all cases.

## Data generating process

The benchmarking is performed using Friedman's function [@friedman1991multivariate] as data-generating process. Friedman's function is augmented with additive group-specific random intercepts with `simulate_friedman_rfx()`/`simulate_friedman_rfx_binary()` (`R/dgp.R`). The random-intercepts are generated with $\sigma^2_G = `r params$sd_group_continuous^2`$ for continuous case, and $\sigma^2_G = `r params$sd_group_binary^2`$ in the binary case.

For the continuous outcome, independent $N(0, \sigma^2)$ noise is added on top of Friedman's function plus the group intercepts^[By default $\sigma = 1$ but this quarto file contains the `y_sd` document parameter to vary this.].

### Signal-to-noise metrics (continuous case)

We consider the identifiability of the mean function and group intercepts. Below we document a signal-to-noise ratio to document the problem difficulty.

```{r snr-demo}
set.seed(1)
n_snr <- 5000
X_snr <- as.data.frame(matrix(runif(n_snr * 10), ncol = 10))
colnames(X_snr) <- paste0("x", 1:10)
var_friedman <- var(friedman_f(X_snr))
```

#### Mean recovery

This treats both the observation noise and the group intercepts as nuisance variance adding difficulty to recovering $f(X)$. The numerator measures Friedman's function's variation across simulated covariates.
$$\text{SNR}_\text{mean} = \frac{\text{Var}(f(X))}{\sigma^2 + \sigma_G^2}.$$ This does not account for $n\in \{`r paste(params$n_values,sep = ", ")`\}$. Smaller $n$ will also increase the problem difficulty.

```{r mean-recovery-demo}
snr_mean <- var_friedman / (params$y_sd^2 + params$sd_group_continuous^2)
```

The estimated variance of Friedman's function is `r round(var_friedman, 2)`, against a combined nuisance variance of $\sigma^2 + \sigma_G^2 = `r round(params$y_sd^2 + params$sd_group_continuous^2, 2)`$, giving a mean-recovery ratio of `r round(snr_mean, 2)`, or **`r round(sqrt(snr_mean), 2)`** on the standardised scale (square root).

#### Intercept recovery

This treats both the observation noise and Friedman's function's variation across covariates as nuisance variance.  $$\text{SNR}_\text{intercept} = \frac{\sigma_G^2}{\text{Var}(f(X)) + \sigma^2}.$$ 


This does not account for $\bar{n} = \frac{n}{n_G} \in \{ `r paste(params$n_values/params$n_groups,sep = ", ")`\}$. Smaller $\bar{n}$ will also increase the problem difficulty.


```{r intercept-recovery-demo}
snr_intercept <- params$sd_group_continuous^2 / (var_friedman + params$y_sd^2)
```

The group-intercept variance is $\sigma_G^2 = `r round(params$sd_group_continuous^2, 2)`$, against a combined nuisance variance of $\text{Var}(f(X)) + \sigma^2 = `r round(var_friedman + params$y_sd^2, 2)`$, giving an intercept-recovery ratio of `r round(snr_intercept, 2)`, or **`r round(sqrt(snr_intercept), 2)`** on the standardised scale (square root).

## Hyperparameters

The main hyperparameters are specified below.

```{r hyperparams}
hp <- baseline_hyperparams()
hp$burn <- hp$burn
hp$draws <- params$n_mcmc_draws
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
hp_vals <- unlist(hp)
hp_display <- ifelse(hp_vals == round(hp_vals), sprintf("%.0f", hp_vals), as.character(hp_vals))
names(hp_display) <- sprintf("`%s`", names(hp))
knitr::kable(t(hp_display), align = "c", row.names = FALSE)
```

For the `baseline` variant, most of the hyperprior parameter specification is trivial to match across packages. However, some hyperpriors are chosen based on the data itself (more or less an empirical Bayes decision).

### Tree-structure prior

Naming conventions for the prior on trees differ across packages: `alpha` = `base` and `beta`=`power`.

These control the prior probability that a node at depth $d$ splits further: $$P(\text{split at depth } d) = \alpha (1+d)^{-\beta}.$$
These are constants, passed unchanged to the correct argument for each package.

### Leaf-shrinkage prior

Each leaf's contribution gets a $N(0, \tau^2)$ prior, with $\tau$ set from `k` (the number of prior standard deviations a leaf is expected to sit from the boundary of $y$'s range) and the number of trees:

$$
\tau = \begin{cases}
\dfrac{\max(y) - \min(y)}{2k\sqrt{\texttt{n\_trees}}} & \text{continuous outcome - uses the data's own range} \\[8pt]
\dfrac{3}{k\sqrt{\texttt{n\_trees}}} & \text{binary outcome - calibrated to the probit latent's } \pm 3 \text{ range instead}
\end{cases}
$$

Some differences between packages:

- `stan4bart` calculates $\tau$ internally from the user-specified `k` argument (as with `dbarts::bart2`).
- `stochtree` requires the specification of $\tau$ directly. It also standardizes continuous outcomes internally, so it requires the  $\tau^2$ to be specified on this standardized scale instead --- we calculate the standardisation manually with $\tau^2 / \text{var}(y)$.

More generally, `stochtree` centres continuous responses before fitting (`stochtree` also rescales it, per the standardisation above), while `stan4bart` work on the raw scale throughout.

### Global error-variance prior {#sec-prior-global-var}

:::{.callout-note}
The following applies to the real-valued response only.
:::

Unlike `BART`/`dbarts` there is no empirical-Bayes style calibration of the prior for the global noise variance in `stochtree` or `stan4bart`. Since we observed such a data-calibrated prior to work well in the vanilla prediction benchmark, we manually construct this calibrated prior for the `baseline` variants. However, only `stochtree` package supports inverse-gamma priors for the global variance. We develop a special matching procedure for `stan4bart` using truncated student-t distributions. The details are given below.


#### Inverse-gamma prior for `stochtree`

Similar to the vanilla prediction `baseline` variant for `stochtree`, we construct the data-calibrated prior of the variance as follows. A crude estimate of the variance is estimated by fitting a linear model against covariates and group-specific random effects (using `lme4::lmer` if it is available). We denote the residual variance estimate from this model as $\hat\sigma^2$.

For `stochtree` the prior needs to be set on the internally-standardized scale it expects for a continuous outcome, and it is specified by with `shape`/`scale` arguments. Overall, this specification is

$$
\lambda = \hat\sigma^2 \cdot \frac{\chi^2_{\nu,\,1-\texttt{sigquant}}}{\nu},
\qquad \nu = \texttt{sigdf}, \qquad
\text{shape} = \frac{\nu}{2}, \qquad \text{scale} = \frac{\nu\lambda}{2\,\text{var}(y)},
$$
where $\chi^2_{\nu,\,p}$ is the $p$-quantile of a $\chi^2_\nu$ distribution.

#### Half-Student-t prior for `stan4bart`

The `stan4bart` package specifies the global **standard deviation** prior via `prior_aux`, a convenience wrapper that can choose from a fixed set of priors (`normal`, `student_t`/`cauchy`, `exponential`, `hs`, `hs_plus`, `laplace`, `lasso`). Given the limited choices it is not possible to match the inverse-gamma distribution (on the variance) exactly.

If $\sigma^2 \sim \text{IG}(\alpha, \beta)$, then $\sigma$ follows a **generalised inverse gamma** distribution ($c = 2$ case) with probability density and cumulative density functions

$$
p(\sigma) = \frac{2\beta^{\alpha}}{\Gamma(\alpha)}\, \sigma^{-2\alpha-1} \exp\!\left(-\frac{\beta}{\sigma^{2}}\right),
\qquad
F(\sigma \le s) = Q\!\left(\alpha,\, \frac{\beta}{s^{2}}\right),
$$
where $Q$ is the regularised upper incomplete gamma function. This is the target we will approximate for the `stan4bart` `baseline` variant. The mode and tail behaviour of the generalised inverse gamma are:

$$
\text{mode} = \sqrt{\frac{2\beta}{2\alpha+1}},
\qquad
p(\sigma) \propto \sigma^{-(\nu+1)} \text{ as } \sigma \to \infty,
$$

with $\alpha = \nu/2$ and $\beta = \nu\lambda/2$. This is the target we would like the `stan4bart` approximation to match as closely as possible.

:::{.callout-note}
## Location of half Student-t

We use a half Student-t prior centered at zero for the `baseline` variant of `stan4bart`. Ideally, we would vary the location parameter of the Student-t and truncate at zero, but it does not seem possible to truncate a Student-t prior that is not centered at zero in  `stan4bart`.

Exploring fixes of this is left for future work.

:::

We calibrate the Student-t distribution (centered at zero) on the remaining two free parameters:

- Degrees of freedom is set to $\nu$, the same value used by every other engine's prior - a $t_\nu$ density decays as $\sigma^{-(\nu+1)}$, matching the generalised inverse gamma tail.
- Scale is chosen so the `sigquant`-quantile of $|T_\nu(0, \texttt{scale})|$ lands exactly on $\hat\sigma$, matching the same calibration principle used for every other engine's global error-variance prior. This has a closed form: for $\sigma = |T|$, $P(\sigma < x) = 2F_\nu(x/\texttt{scale}) - 1$ where $F_\nu$ is the $t_\nu$ CDF, so
$$
\texttt{sigquant} = 2F_\nu\!\left(\hat\sigma/\texttt{scale}\right) - 1
\;\Longrightarrow\;
\texttt{scale} = \frac{\hat\sigma}{F_\nu^{-1}\!\left(\dfrac{1+\texttt{sigquant}}{2}\right)}.
$$

As the mode is fixed at $0$ rather than matched to the target, this prior cannot reproduce the generalised inverse gamma's *shape* (interior peak) - only its tail weight ($\nu$) and one quantile ($\hat\sigma$) carry over. The comparison plot below makes the resulting shape difference visible directly.

#### Comparison of priors

The `default` variant, which uses each package's own out-of-the-box choice:

- data-calibrated exponential prior for `stan4bart` on the standard deviation, and 
- `stochtree`'s default inverse-gamma is a $\text{IG}(0,0)$ distribution (an improper Jeffery's prior) on the variance.

The `stan4bart` default is

$$
\sigma \sim \text{Exponential}\!\left(\text{rate} = 1/\text{sd}(y)\right),
$$

and the `default+sigest` variant keeps this same exponential family but recalibrates the rate from $\hat\sigma$ instead of the raw $\text{sd}(y)$,

$$
\sigma \sim \text{Exponential}\!\left(\text{rate} = 1/\hat\sigma\right),
$$

accounting for the presence of subject-specific intercepts.

`stochtree`'s default is an improper prior with density $p(\sigma^2) \propto \frac{1}{\sigma^2}$ on the variance, or equivalently

$$
p(\sigma) \propto \frac{1}{\sigma},
$$

on the standard deviation.

The following comparison plots these an example of these priors (minus the improper prior) on the standard deviation scale. The data-calibrated priors are constructed by simulating $n = `r min(params$n_values)`$ samples and using equivalent settings to the simulation to be performed.

```{r prior-compare-plot}
#| fig-height: 4
#| fig-cap: "Priors on $\\sigma$ (using from a single simulated dataset for calibration at the smallest n). stochtree's default is improper and is not shown. The dashed line marks `sigest`, the dotted line the true $\\sigma$."
sigma_gig_density <- function(y, alpha, beta) {
  2 * beta^alpha / gamma(alpha) * y^(-2 * alpha - 1) * exp(-beta / y^2)
}
sigma_trunc_t_density <- function(x, df, location, scale) {
  ifelse(x <= 0, 0,
    dt((x - location) / scale, df = df) / scale / (1 - pt(-location / scale, df = df))
  )
}

# One dataset, same DGP settings and same n as the benchmark's smallest run.
set.seed(params$seed)
sim_prior <- simulate_friedman_rfx(
  n = min(params$n_values), n_groups = params$n_groups,
  sd_group = params$sd_group_continuous, y_sd = params$y_sd
)
X_prior <- sim_prior$data[, setdiff(names(sim_prior$data), c("y", "group"))]
y_prior <- sim_prior$data$y
group_prior <- sim_prior$data$group

# sigest via the engines' own helper, group included so the intercepts are not
# absorbed into the residual; then the inverse-gamma target it implies.
sigest_demo <- compute_sigest(X_prior, y_prior, group = group_prior)
sigest_nogroup_demo <- compute_sigest(X_prior, y_prior)
nu <- hp$sigdf
lambda_demo <- sigest_demo^2 * qchisq(1 - hp$sigquant, df = nu) / nu
alpha_demo <- nu / 2
beta_demo <- nu * lambda_demo / 2

# The half-Student-t actually passed to stan4bart, taken from the engine's own helper.
prior_aux_demo <- hp_to_stan4bart_sigma_prior(hp, X_prior, y_prior, group = group_prior)$prior_aux
# stan4bart's default: exponential(rate = 1, autoscale = TRUE) -> rate = 1/sd(y).
exp_rate_demo <- 1 / sd(y_prior)
# default+sigest: same exponential family, rate = 1/sigest instead of 1/sd(y).
exp_sigest_demo <- hp_to_stan4bart_sigma_prior_exp_sigest(hp, X_prior, y_prior, group = group_prior)$prior_aux

sigma_grid <- seq(1e-4, 3 * sigest_demo, length.out = 600)

# stochtree's own default is omitted: it is the improper Jeffreys prior
# (see the note below the plot), which has no density to draw.
prior_curves <- bind_rows(
  tibble(x = sigma_grid, density = sigma_gig_density(sigma_grid, alpha_demo, beta_demo),
         prior = "stochtree baseline (generalised inverse gamma)"),
  tibble(x = sigma_grid, density = sigma_trunc_t_density(sigma_grid, prior_aux_demo$df,
                                                          prior_aux_demo$location, prior_aux_demo$scale),
         prior = "stan4bart baseline (half-Student-t)"),
  tibble(x = sigma_grid, density = dexp(sigma_grid, rate = exp_rate_demo),
         prior = "stan4bart default (exponential)"),
  tibble(x = sigma_grid, density = dexp(sigma_grid, rate = 1 / exp_sigest_demo$scale),
         prior = "stan4bart default+sigest (exponential)")
)

prior_curves %>%
  ggplot(aes(x = x, y = density, colour = prior)) +
  geom_line() +
  geom_vline(xintercept = sigest_demo, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = params$y_sd, linetype = "dotted", colour = "grey30") +
  labs(x = expression(sigma), y = "Density", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 1))
```

At $n = `r min(params$n_values)`$ this gives $\hat\sigma = `r round(sigest_demo, 2)`$ against a true $\sigma = `r params$y_sd`$. 

:::{.callout-warning}
## Future work

For $n=250$ (at least), even with $\hat\sigma^2$ overestimated, the bulk of the IG prior mass lines up suspiciously well with the truth. Prior sensitivity should be tested in a future benchmark.
:::

It doesn't seem possible with `stan4bart`'s existing priors to replicate the internal peak of the calibrated inverse-gamma used by `stochtree`.  

## Run

<details>

```{r run-benchmark}
n_values <- if (isTRUE(params$fast_mode)) c(60, 100) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps
n_for_examples <- max(n_values)

cache_file <- file.path("cache", sprintf("rfx_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  old_plan <- future::plan(future::multisession, workers = benchmark_worker_count(params$parallel_workers))
  on.exit(future::plan(old_plan), add = TRUE)
  run_benchmark_rfx(n_values = n_values, B = n_reps, hp = hp, seed = params$seed, y_sd = params$y_sd,
                     sd_group_continuous = params$sd_group_continuous, sd_group_binary = params$sd_group_binary,
                     n_groups = params$n_groups,
                     mcmc_seed = params$mcmc_seed, benchmark_dir = benchmark_dir)
})
```

</details>

## Results

### Prediction accuracy

`fit_time_sec` is reported as a median, not a mean, to prevent (potential) extreme outliers affecting the result (check @sec-runtime) .

```{r summary-table}
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(
    across(c(rmse, mae, coverage95, crps), mean),
    fit_time_sec = median(fit_time_sec),
    .groups = "drop"
  ) %>%
  select(outcome, n, engine, variant, rmse, mae, coverage95, crps, fit_time_sec) %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3, format = "html") %>%
  kableExtra::collapse_rows(columns = 1:3, valign = "top")
```

### Predicted vs true

These plots show one representative replication for `` n = `r n_for_examples` `` (the largest value tested).

```{r pred-scatter-continuous}
#| fig-height: 4
res$examples %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 2) +
  labs(title = "Continuous outcome", x = "True value (f(x) + group effect)", y = "Posterior mean fitted value") +
  theme_bw()
```

```{r pred-scatter-binary}
#| fig-height: 4
res$examples %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 2) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Binary outcome (probability scale)", x = "True probability", y = "Posterior mean fitted probability") +
  theme_bw()
```

### Group-effect recovery

Here we look at the recovery of the group-specific intercepts. The trees in a BART model can absorb part of the overall baseline level that would otherwise be captured in the random intercepts. That is, the group mean can be split between "trees" and "random intercepts", so the raw random intercepts may not exactly match the true random intercepts. 

To account for this we demean the intercept estimates and true values before we compare. Define $\bar{\hat u} = \frac{1}{n_G}\sum_g \hat u_g$ and $\bar u = \frac{1}{n_G}\sum_g u_g$, we consider the correlation:

$$
\texttt{group\_effect\_cor} = \frac{\sum_g (\hat u_g - \bar{\hat u})(u_g - \bar u)}{\sqrt{\sum_g (\hat u_g - \bar{\hat u})^2} \, \sqrt{\sum_g (u_g - \bar u)^2}}
$$

which is the ordinary Pearson correlation across groups (shift- and scale-invariant). It checks whether groups with a larger true effect also get a larger estimated one, in the right order. We also consider the centered RMSE:

$$
\texttt{group\_effect\_centered\_rmse} = \sqrt{\frac{1}{n_G} \sum_g \Big[(\hat u_g - \bar{\hat u}) - (u_g - \bar u)\Big]^2}
$$

which represents an actual recovery test (excluding the shift). A third, uncentred version is reported alongside them for reference:

$$
\texttt{group\_effect\_raw\_rmse} = \sqrt{\frac{1}{n_G} \sum_g \big(\hat u_g - u_g\big)^2}
$$

but isn't meaningful if used in isolation.

```{r group-effect-table}
res$group_metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(across(c(group_effect_cor, group_effect_centered_rmse, group_effect_raw_rmse), mean), .groups = "drop") %>%
  select(outcome, n, engine, variant, group_effect_cor, group_effect_centered_rmse, group_effect_raw_rmse) %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3, format = "html") %>%
  kableExtra::collapse_rows(columns = 1:3, valign = "top")
```


The scatter plots below are centred the same way as the centred RMSE (see the axis labels).

```{r group-effect-scatter-prep}
# Centred, matching group_effect_recovery()'s own metric definition: the
# shared level between BART's trees and the random intercept isn't
# identified, so only the *relative* effects are comparable.
group_examples_centred <- res$group_examples %>%
  group_by(outcome, engine, variant) %>%
  mutate(
    true_centred = true_group_effect - mean(true_group_effect),
    hat_centred = group_effect_hat - mean(group_effect_hat)
  ) %>%
  ungroup()
```

```{r group-effect-scatter-continuous}
#| fig-height: 4
group_examples_centred %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = true_centred, y = hat_centred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 1.5) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Continuous outcome", x = "True group effect (centred)", y = "Posterior mean group effect estimate (centred)") +
  theme_bw()
```

```{r group-effect-scatter-binary}
#| fig-height: 4
group_examples_centred %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = true_centred, y = hat_centred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 1.5) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Binary outcome (probit latent scale)", x = "True group effect (centred)", y = "Posterior mean group effect estimate (centred)") +
  theme_bw()
```

### Cross-engine agreement (baseline rows)

Do `stan4bart` and `stochtree::bart` agree with each other on the same simulated dataset, independent of the truth? Only `baseline` rows are compared, for the same reason as `benchmark-prediction.qmd`'s equivalent section: `default` and `baseline+gfr` use different priors or warm-start settings, which would muddy a same-prior comparison.

```{r agreement-table}
res$agreement %>%
  group_by(outcome, engine_a, engine_b) %>%
  summarise(mean_correlation = mean(correlation), mean_rmse = mean(rmse), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

### Coverage: nominal vs empirical

```{r coverage-plot}
#| fig-height: 4
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(coverage95 = mean(coverage95), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = coverage95, colour = factor(n))) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### MCMC mixing {#sec-mcmc-mixing}

Here we take a look at ESS of $\sigma$ (global standard deviation) and $\sigma_G$ (random effect standard deviation) across the engines, variants, and settings. Then the autocorrelation plots. Later we look at the quantiles of the estiamted fit.

```{r ess-table}
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(across(c(sigma_global_ess, sigma_group_ess), ~if (all(is.na(.x))) NA_real_ else mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 0, format = "html") %>%
  kableExtra::collapse_rows(columns = 1:3, valign = "top")
```

```{r ess-plot}
#| fig-height: 5
res$metrics %>%
  select(outcome, engine, variant, n, sigma_global_ess, sigma_group_ess) %>%
  tidyr::pivot_longer(c(sigma_global_ess, sigma_group_ess), names_to = "parameter", values_to = "ess") %>%
  filter(!is.na(ess)) %>%
  mutate(parameter = recode(parameter, sigma_global_ess = "sigma (global)", sigma_group_ess = "sigma_G (random intercept)")) %>%
  ggplot(aes(x = paste(engine, variant), y = ess, colour = factor(n))) +
  geom_boxplot(outlier.alpha = 0.5) +
  facet_grid(outcome ~ parameter) +
  coord_flip() +
  labs(x = NULL, y = "Effective sample size", colour = "n") +
  theme_bw()
```

A low or engine-specific ESS here is a flag to interpret RMSE/coverage differences with some caution - a noisier posterior mean widens the gap between engines without either one's model or prior actually being worse.

#### Autocorrelation of $\sigma$

```{r acf-plot}
#| fig-height: 6
res$acf_global %>%
  filter(outcome == "continuous", lag > 0) %>%
  group_by(engine, variant, n, lag) %>%
  summarise(acf = mean(acf), .groups = "drop") %>%
  ggplot(aes(x = lag, y = acf, colour = factor(n))) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line() +
  facet_wrap(~ paste(engine, variant)) +
  labs(x = "Lag (raw MCMC sweeps)", y = expression(paste("Autocorrelation of ", sigma)), colour = "n") +
  theme_bw()
```

#### Do the trees mix at the same rate?

Both engines update their residual-variance parameter conditional on the *current* tree ensemble every sweep. `stochtree` updates the random effects and variance terms each in conditional/conjugate Gibbs steps, whilst `stan4bart` uses a Stan HMC step for all non-Bart parameters. Here five quantiles $(0.05, 0.2, 0.5, 0.8, 0.95)$ of the fitted values across observations are tracked, computed per retained draw.

```{r fitted-quantile-acf-plot}
#| fig-height: 8
res$fitted_quantile_acf %>%
  filter(lag > 0) %>%
  mutate(quantile = factor(quantile, levels = c("q5", "q20", "q50", "q80", "q95"))) %>%
  group_by(engine, variant, n, quantile, lag) %>%
  summarise(acf = mean(acf), .groups = "drop") %>%
  ggplot(aes(x = lag, y = acf, colour = paste(engine, variant))) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line() +
  facet_grid(quantile ~ n, labeller = label_both) +
  labs(x = "Lag (raw MCMC sweeps)", y = "Autocorrelation of fitted-value quantile", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 2))
```


```{r fitted-quantile-correlation-table}
res$fitted_quantile_joint %>%
  filter(outcome == "continuous") %>%
  group_by(engine, variant) %>%
  summarise(
    `cor(q95, sigma)` = cor(q95, sigma),
    `cor(q95, sigma_G)` = cor(q95, sigma_G),
    `cor(sigma, sigma_G)` = cor(sigma, sigma_G),
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 2)
```


```{r fitted-quantile-pairs-setup}
pairs_data <- function(engine_name, variant_name) {
  res$fitted_quantile_joint %>%
    filter(outcome == "continuous", engine == engine_name, variant == variant_name) %>%
    transmute(sigma_G = sigma_G, sigma = sigma, q5 = q5, q95 = q95)
}

pairs_plot <- function(df, title) {
  GGally::ggpairs(
    df,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 1)),
    diag = list(continuous = "densityDiag"),
    upper = list(continuous = "blank"),
    title = title
  ) + theme_bw()
}
```

```{r fitted-quantile-pairs-stan4bart}
#| fig-height: 6
#| fig-width: 6
pairs_plot(pairs_data("stan4bart", "baseline"), "stan4bart (baseline)")
```

```{r fitted-quantile-pairs-stochtree-baseline}
#| fig-height: 6
#| fig-width: 6
pairs_plot(pairs_data("stochtree::bart", "baseline"), "stochtree::bart (baseline)")
```

### Runtime {#sec-runtime}

The box plot below shows the full per-replication fit-time distribution on a
log scale, which is where the outliers actually show up - as individual
points well above the whiskers, rather than as a shift in a single summary
number.

```{r runtime-boxplot}
#| fig-height: 4
res$metrics %>%
  ggplot(aes(x = paste(engine, variant), y = fit_time_sec, colour = factor(n))) +
  geom_boxplot(outlier.alpha = 0.5) +
  facet_wrap(~outcome) +
  coord_flip() +
  scale_y_log10() +
  labs(x = NULL, y = "Fit time (s, log scale)", colour = "n") +
  theme_bw()
```

### Flags

This section lists rows whose RMSE (for prediction) or centred group-effect
RMSE (for recovery) is a statistical outlier (|z| > 2) within its own
outcome/`n` setting. See `flag_outliers()` in `R/metrics.R` for the caveat
about small-group z-scores - with only four rows per group here, this
flagging is necessarily a coarse screen, not a precise test.

```{r flags}
flagged_rmse <- res$metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(rmse = mean(rmse), .groups = "drop") %>%
  flag_outliers("rmse", c("outcome", "n"))

flagged_group <- res$group_metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(group_effect_centered_rmse = mean(group_effect_centered_rmse), .groups = "drop") %>%
  flag_outliers("group_effect_centered_rmse", c("outcome", "n"))

if (nrow(flagged_rmse) == 0) {
  cat("No prediction RMSE outliers flagged.\n")
} else {
  knitr::kable(flagged_rmse, digits = 3, caption = "Prediction RMSE outliers")
}

if (nrow(flagged_group) == 0) {
  cat("No group-effect-recovery outliers flagged.\n")
} else {
  knitr::kable(flagged_group, digits = 3, caption = "Group-effect recovery outliers")
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
