---
title: "Benchmark: Prediction With Random Effects"
subtitle: "stan4bart vs. stochtree::bart(rfx_group_ids), on ground both were built for"
format:
  html:
    toc: true
    code-fold: true
params:
  fast_mode: true
  n_values: [250, 1000]
  n_reps: 20
  seed: 1
---

```{r setup}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytreatment)
library(tidybayes)
library(stan4bart)
library(stochtree)
library(dbarts)

source("R/cache.R")
source("R/hyperparams.R")
source("R/dgp.R")
source("R/metrics.R")
source("R/fit-engines-rfx.R")
source("R/metrics-rfx.R")
source("R/run-benchmark-rfx.R")

sessioninfo::session_info()
```

## Scope

`benchmark-prediction.qmd` excludes `stan4bart` entirely, because it is
architecturally a mixed-effects-plus-BART model: its formula parser requires
at least one `(1|group)` random-effects term, so there is no fair "no
groups" version of it to test there. This document gives it the ground it
was actually built for instead, comparing it against `stochtree::bart()` fit
with `rfx_group_ids_train` - intercept-only random effects, the direct
analogue of a `(1|group)` term. `BART`, `bartCause`, and `stochtree::bcf`
have no random-effects prediction mode of their own, so they are excluded
here. `bartMachine` is excluded throughout, as in the other benchmark
documents.

There are four rows in total: `stan4bart` at baseline and with `+auto_k`,
and `stochtree::bart` at baseline and with `baseline+gfr`.

The data-generating process is
`simulate_friedman_rfx()`/`simulate_friedman_rfx_binary()` (`R/dgp.R`):
Friedman's function plus an explicit random intercept for each group. Unlike
`simulate_su_hill_data()`'s own `n_subjects`/`sd_subjects` mechanism - which
adds the random effect straight into the noisy `y` and never returns the
true effects at all - this DGP hands back exact ground truth for both the
fixed-effect surface and the true per-group effects.

Beyond the usual prediction-accuracy metrics, the sharpest diagnostic in
this document is **group-effect recovery**: does each engine's own posterior
mean per-group random intercept actually track the true simulated group
effects? One wrinkle worth keeping in mind when reading it: BART's trees can
absorb some of the overall baseline level that would otherwise land in the
random intercepts, since the two are not uniquely identified from the data
alone. For that reason, recovery is reported as a correlation plus a
*centred* RMSE (with any shared shift removed) rather than as a raw bias.

## Hyperparameters

```{r hyperparams}
hp <- baseline_hyperparams()
if (isTRUE(params$fast_mode)) {
  hp$burn <- 20
  hp$draws <- 30
  hp$n_trees <- 20
}
str(hp)
```

## Run

<details>
<summary>Show run log</summary>

```{r run-benchmark}
n_values <- if (isTRUE(params$fast_mode)) c(60, 100) else params$n_values
n_reps <- if (isTRUE(params$fast_mode)) 2 else params$n_reps

cache_file <- file.path("cache", sprintf("rfx_%s.rds", if (isTRUE(params$fast_mode)) "fast" else "full"))
res <- cached_run(cache_file, function() {
  run_benchmark_rfx(n_values = n_values, B = n_reps, hp = hp, seed = params$seed)
})
```

</details>

## Results

### Prediction accuracy

```{r summary-table}
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(across(c(rmse, mae, coverage95, crps, fit_time_sec), mean), .groups = "drop") %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

### Predicted vs. true (fixed-effect surface + group effect combined)

As with the other benchmark documents, this plot shows one representative
replication - the one with the largest `n`.

```{r pred-scatter-continuous}
#| fig-height: 4
res$examples %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Continuous outcome", x = "True value (f(x) + group effect)", y = "Posterior mean fitted value") +
  theme_bw()
```

```{r pred-scatter-binary}
#| fig-height: 4
res$examples %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = truth, y = fitted)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Binary outcome (probability scale)", x = "True probability", y = "Posterior mean fitted probability") +
  theme_bw()
```

### Group-effect recovery

This is the headline diagnostic for the whole document: does each engine's
own posterior mean per-group random intercept track the true simulated group
effects? The values plotted are centred (see the axis labels), which makes
them comparable in scale across engines within the same outcome - so each
outcome gets a single plot with shared, not free, axes across its engine
panels. Giving the engine dimension its own free scale via
`facet_grid(..., scales = "free")` would let each panel float on its own
axis range independently, which would be misleading for exactly this
side-by-side comparison - the same issue that is diagnosed in more detail
for the predicted-vs-true plot in `benchmark-prediction.qmd`.

```{r group-effect-table}
res$group_metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(across(c(group_effect_cor, group_effect_centered_rmse), mean), .groups = "drop") %>%
  arrange(outcome, n, engine, variant) %>%
  knitr::kable(digits = 3)
```

```{r group-effect-scatter-prep}
# Centred, matching group_effect_recovery()'s own metric definition: the
# shared level between BART's trees and the random intercept isn't
# identified, so only the *relative* effects are comparable.
group_examples_centred <- res$group_examples %>%
  group_by(outcome, engine, variant) %>%
  mutate(
    true_centred = true_group_effect - mean(true_group_effect),
    hat_centred = group_effect_hat - mean(group_effect_hat)
  ) %>%
  ungroup()
```

```{r group-effect-scatter-continuous}
#| fig-height: 4
group_examples_centred %>%
  filter(outcome == "continuous") %>%
  ggplot(aes(x = true_centred, y = hat_centred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 1.5) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Continuous outcome", x = "True group effect (centred)", y = "Posterior mean group effect estimate (centred)") +
  theme_bw()
```

```{r group-effect-scatter-binary}
#| fig-height: 4
group_examples_centred %>%
  filter(outcome == "binary") %>%
  ggplot(aes(x = true_centred, y = hat_centred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 1.5) +
  facet_wrap(~ paste(engine, variant), nrow = 1) +
  labs(title = "Binary outcome (probit latent scale)", x = "True group effect (centred)", y = "Posterior mean group effect estimate (centred)") +
  theme_bw()
```

### Coverage: nominal vs. empirical

```{r coverage-plot}
#| fig-height: 4
res$metrics %>%
  group_by(outcome, engine, variant, n) %>%
  summarise(coverage95 = mean(coverage95), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = coverage95, colour = factor(n))) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  facet_wrap(~outcome) +
  coord_flip() +
  labs(x = NULL, y = "Empirical 95% interval coverage", colour = "n") +
  theme_bw()
```

### Runtime

```{r runtime-plot}
#| fig-height: 3
res$metrics %>%
  group_by(outcome, engine, variant) %>%
  summarise(fit_time_sec = mean(fit_time_sec), .groups = "drop") %>%
  ggplot(aes(x = paste(engine, variant), y = fit_time_sec, fill = outcome)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Mean fit time (s)") +
  theme_bw()
```

### Flags

This section lists rows whose RMSE (for prediction) or centred group-effect
RMSE (for recovery) is a statistical outlier (|z| > 2) within its own
outcome/`n` setting. See `flag_outliers()` in `R/metrics.R` for the caveat
about small-group z-scores - with only four rows per group here, this
flagging is necessarily a coarse screen, not a precise test.

```{r flags}
flagged_rmse <- res$metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(rmse = mean(rmse), .groups = "drop") %>%
  flag_outliers("rmse", c("outcome", "n"))

flagged_group <- res$group_metrics %>%
  group_by(outcome, n, engine, variant) %>%
  summarise(group_effect_centered_rmse = mean(group_effect_centered_rmse), .groups = "drop") %>%
  flag_outliers("group_effect_centered_rmse", c("outcome", "n"))

if (nrow(flagged_rmse) == 0) {
  cat("No prediction RMSE outliers flagged.\n")
} else {
  knitr::kable(flagged_rmse, digits = 3, caption = "Prediction RMSE outliers")
}

if (nrow(flagged_group) == 0) {
  cat("No group-effect-recovery outliers flagged.\n")
} else {
  knitr::kable(flagged_group, digits = 3, caption = "Group-effect recovery outliers")
}
```

## Session info

```{r session-info-end}
sessioninfo::session_info()
```
